Examining spatial equity and accessibility to a public bicycle share program using a balanced floating catchment area approach

### 2021-03-24 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Manuscript draft completed and sent to co-authors for review

This paper uses the city of Hamilton, located in Ontario, Canada, as a case study. The city launched a public bicycle share program, Hamilton Bike Share, in March 2015 with 115 stations and 750 bicycles [@hamiltonHamiltonBikeShare2015]. Before June 2020, the program was known as SoBi Hamilton. Stations are spaced between 300 and 600 metres apart [@scottRouteChoiceBike2021]. The core service area spans 35 km^2 of the city and roughly 138,000 can reach a bike share hub within 30 minutes of walking [see Figure 1]. This represents roughly one fifth of the population in Hamilton, according to the 2016 Canadian Census. The program was enthusiastically welcomed in the city - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015]. In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, Everyone Rides Initiative (ERI), to remove barriers that may prevent individuals from accessing bike share in Hamilton. An additional 75 bicycles and 15 stations were added to the program, which expanded it to more disadvantaged areas in the core service area [see Figure 2]. The program also offers subsidized memberships to individuals who identify as low income. A comparable program can be found in Philadelphia [see @caspiBikesharingPhiladelphiaLowerincome2019]. As of June 2020, the bike share program has 900 bikes and 130 stations, and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].

@@

```{r hamilton-and-sobi-service-area, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "The core service area of Hamilton Bike Share is outlined in blue. Dissemination areas in the Hamilton Census Metropolitan Area (CMA) are outlined in grey."}
ggplot() + 
  geom_sf(data = hamilton_da_2016,
          color = NA,
          fill = "gray") + 
  geom_sf(data = sobi_service,
          fill = NA,
          colour = "black") + 
  theme(panel.background = element_rect(fill = "white"))
```

```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "100%", fig.cap = "The spatial distribution of bike share stations in Hamilton, Ontario. The service area of Hamilton Bike Share is outlined in blue and the city's downtown core is outlined in light blue. Dissemination areas within the Hamilton CMA that touch the bounding box of the core service area are outlined in grey."}
ggplot() + 
  geom_sf(data = cropped_da,
          color = NA,
          fill = "gray") +
  geom_sf(data = sobi_service, 
          fill = NA,
          color = "black") +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          size = 1,
          color = "dark green") +
  geom_sf(data = sobi_hubs %>% 
            dplyr::filter(hub_type == "Conventional" & 
                            hub_status == "Active"), # Remove deactivated hubs
          aes(size = RACKS_AMOU),
          shape = 21,
          fill = "blue") +
  geom_sf(data =  sobi_hubs %>% 
            dplyr::filter(hub_type == "ERI"), 
          aes(size = RACKS_AMOU), 
          shape = 21,
          fill = "green") +
  scale_size(name = "Number of racks", range = c(0.5, 4)) +
  theme(panel.background = element_rect(fill = "white"))
```

@@

We first plotted the population in all dissemination areas within SoBi Hamilton's core service area. Since pycnophylactic interpolation occurs at such a micro scale, we had to ensure that population numbers were not interpolated or allocated to areas where people do not live in Hamilton (i.e., where schools or parks are located). To do so, we retrieved shapefiles for various geographic features (see Table 1) from Open Hamilton. Figure 3 shows where these spaces are located within SoBi Hamilton's core service area. Next, we extracted these features from the PBSP core service area and used pycnophylactic interpolation to disaggregate and reallocate population from dissemination areas to smaller polygons. These micro-level polygons are 50 by 50 metres.

@@

kable(data_tbl,
      "latex",
      booktabs = TRUE,
      caption = "\\label{tab:landscape-features}Landscape features") %>%
  kable_paper(full_width = F) %>%
  column_spec(1, bold = T, border_right = T) %>%
  column_spec(2, width = "30em") %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

@@

```{r figure-2, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Population in all dissemination areas in the SoBi Hamilton core service area and within 30 minutes of walking to a bike share station."}
ggplot() +
  geom_sf(data = hamilton_da_2016 %>%
            dplyr::filter(sobi_service == "Yes"),
          aes(fill = population),
          color = "white") +
  geom_sf(data = sobi_service,
          fill = NA) +
  scale_fill_distiller(palette = "YlOrRd", 
                       direction = 1) +
  theme(panel.background = element_rect(fill = "white"))
```

```{r figure-4, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Interpolated population cells in the SoBi Hamilton core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue). The downtown area is outlined in dark green. Geographic features have been extracted. Dissemination area polygons are outlined in grey."}
ggplot() + 
  geom_sf(data = cropped_da,
          fill = "gray",
          color = NA) + 
  geom_sf(data = population_50x50,
          aes(color = population)) +
  scale_color_distiller(palette = "YlOrRd", 
                        direction = 1) +
  geom_sf(data = sobi_service,
          fill = NA,
          colour = "black") +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  theme(panel.background = element_rect(fill = "white"))
```

<!-- Suggest to remove this image: all ithe information it gives is already found in other figures: the location of the SoBi stations and the interpolated popualtion

```{r figure-5, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Interpolated population cells in the SoBi Hamilton core service area and within 30 minutes of walking to the core service area. Green circles represent the location of the equity stations that were added in 2018. Blue circles represent the conventional stations. The service area of the PBSP is outlined in black, and dissemination areas from the 2016 Canadian Census are outlined in light gray."}
ggplot() + 
  geom_sf(data = cropped_da, 
          color = NA, 
          fill = "gray") + 
  geom_sf(data = population_50x50, 
          aes(color = population)) +
  geom_sf(data = sobi_hubs %>% 
            dplyr::filter(hub_type == "Conventional" &
                            hub_status == "Active"), 
          size = 1.5, 
          shape = 21, 
          fill = "blue") +
  geom_sf(data =  sobi_hubs %>% 
            dplyr::filter(hub_type == "ERI"), 
          size = 1.5, shape = 21, 
          fill = "green") +
  geom_sf(data = sobi_service, 
          fill = NA, 
          colour = "black") +
  scale_color_distiller(palette = "YlOrRd", 
                        direction = 1) +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  facet_grid(rows = vars(hub_type)) +
  theme(panel.background = element_rect(fill = "white"))
```
--->

@@

Consensus regarding the distance that individuals are willing to walk to access a PBSP is lacking, therefore the literature on walking behaviour was consulted to determine the thresholds for the sensitivity analysis. Previous studies have found that living within 250 metres [@fullerUseNewPublic2011] and 300 metres [@kabraBikeShareSystemsAccessibility2020] is correlated with bike share use, while other research has found that walking trips are less than 600 metres and rarely more than 1200 metres [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 metres [@larsenQuarterMileReexamining2010]. As seen in Table \ref{tab:accessibility-sensibility}, we find that accessibility increases with a threshold between two and four minutes, but is then maximized at 5 minutes. At some stations, Hamilton Bike Share will depict a map to show the user the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are anticipated to walk. Accessibility decreases significantly after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each bike share station.

@@

```{r accessibility-sensibility, eval=FALSE, include = FALSE}
# Testing different thresholds to check how accessibility changes
accessibility_test <- data.frame(threshold = c(3:15), 
                                 accessibility = numeric(13))
for(i in 1:13){
  results_test <- ttm_walk %>%
    dplyr::filter(hub_type == "Conventional" &
                    hub_status == "Active") %>%
    b2sfca(threshold = i +2 )
  accessibility_test$accessibility[i] <- sum(results_test$accessibility$accessibility)
}

@@

@@

At this threshold, we find that there are 25.2 bicycle racks per person in the original system.  With the addition of equity stations, there are now 25.4 bicycles per person. Figure \ref{fig:figure-6} presents a comparison of accessibility between the systems. Accessibility is fairly uniform, with the exception of two small areas near the university and the waterfront park where accessibility is slightly higher. Equity stations are shown in green, which helps to demonstrate how they increase accessibility between the systems.

@@

At this threshold, we find that there are 68.6 bicycle racks per person in the original system.  With the addition of equity stations, there are now 68.8 bicycles per person. Figure \ref{fig:figure-7} presents a comparison of accessibility between the systems. Again, accessibility is fairly uniform, with the exception of one area near the waterfront park where accessibility is higher.

@@

At this threshold, we find that there are 3.61 bicycle racks per person in the original system.  With the addition of equity stations, there are now 3.74 bicycles per person. Figure \ref{fig:figure-8} presents a comparison of accessibility between the systems. We begin to see differences in accessibility across the service area, with users near the university and its adjacent neighbourhoods, as well as neighbourhoods north of the downtown area have slightly higher accessibility. While the differences are modest, they are more apparent at this threshold than at shorter walking distances.

@@

At this threshold, we find that there are 2.44 bicycle racks per person in the original system.  With the addition of equity stations, there are now 2.55 bicycles per person. Figure \ref{fig:figure-9} presents a comparison of accessibility between the systems. Users near the university and the neighbourhoods north of the downtown area (outlined in green) have the highest accessibility, followed by those who live in the city's downtown area. Accessibility in the east end of the core service area remains low.

@@

Approximately 138,000 people live within Hamilton Bike Share's core service area or within a 10 minute walk to the service area. We find that over 118,000 people can access a bicycle share station within a 15 minute walk, which represents roughly 85% of the total population in Hamilton Bike Share's core service area [see Table \ref{tab:table-2}]. While previous research found that neighbourhoods with more disadvantage are better serviced by Hamilton Bike Share [@hosfordWhoArePublic2018], the authors used the Pampalon Deprivation Index from 2011 as a measure of socioeconomic status. Instead, we use median total income for each dissemination area to examine whether disparities in accessibility exist within the core service area. We divide income by quantiles: bottom 20%, second 20%, third 20%, fourth 20%, and top 20%.  One of the unique properties of the balanced floating catchment area method is that data can be reaggregated from small to larger polygons while preserving the total population and number of bicycle racks at each station. This avoids demand and supply inflation, and also enables us to present findings in a way that is easier to interpret (i.e., at the dissemination area level).

Panels 1 and 2 depict bivariate choropleth maps with the combined spatial distribution of accessibility and median total income per household for the different thresholds. Our analysis demonstrates that the stations added to Hamilton's public bicycle share program to increase equity for disadvantaged neighbourhoods achieved their goal by increasing accessibility, albeit only modestly. By implementing equity stations in more areas with lower median total income in Hamilton, the PBSP has achieved greater horizontal equity by extending the spatial distribution of bicycles across the city. This is particularly evident at shorter walking distances of three and five minutes.

However, we found that dissemination areas with lower total household income do not necessarily have greater access to the program [see Table \ref{tab:table-3}]. Income disparities still persist, however only at certain thresholds. With and without equity stations, people in the top 20% of income have the highest level of accessibility at a threshold of ten and fifteen minutes. At distance thresholds of three and five minutes, people in the second 20% have the highest level of accessibility by a significant amount. This is due to the concentration of stations near or within the city's downtown area and the "lower city" which historically has been more disadvantaged compared to the surrounding suburban neighbourhoods. This is an artifact of the city's steel industry which led to more blue collar and working class people living in the city's downtown area and "lower city". However, the bottom 20%, who may benefit the most from Hamilton Bike Share's equity initiatives, have the lowest accessibility at three minutes threshold and the second lowest accessibility at all other thresholds.

On the whole, we find that areas that have less than their equitable share of the level of service are some of the most socioeconomically disadvantaged areas in the city. This finding aligns with other studies that have found disparities in station or bike location along income lines in New York City [@babagoliExploringHealthSpatial2019], Tampa [@chenExploringEquityPerformance2019], and Seattle [@mooneyFreedomStationSpatial2019].

@@

kable(accessibility_income,
      "latex",
      booktabs = TRUE,
      col.names = c("Income Quintile",
                    "Accessibility",
                    "Population",
                    "Accessibility",
                    "Population",
                    "Accessibility",
                    "Population"),
      align = c("l", "c", "c", "c", "c", "c", "c"),
      caption = "\\label{tab:accessibility-by-income}Accessibility and population serviced by income quintile") %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  pack_rows("Threshold - 3 minutes", 1, 5, 
            label_row_css = "background-color: #8f9498; color: #fff;") %>%
  pack_rows("Threshold - 5 minutes", 6, 10, 
            label_row_css = "background-color: #8f9498; color: #fff;") %>%
  pack_rows("Threshold - 10 minutes", 11, 15, 
            label_row_css = "background-color: #8f9498; color: #fff;") %>%
  pack_rows("Threshold - 15 minutes", 16, 20, 
            label_row_css = "background-color: #8f9498; color: #fff;") %>%
  add_header_above(c(" " = 1, "Without Equity Stations" = 2, "With Equity Stations" = 2, "Difference" = 2)) %>%
  #column_spec(2, color = "white",
  #background = spec_color(accessibility_income$accessibility[1:40], end = 1)) %>%
  footnote(general = " ",
           alphabet = c("With equity stations = Hamilton Bike Share current system (118 conventional stations, 12 equity stations)", "Without equity stations = Hamilton Bike Share original system (118 conventional stations, no equity stations)"))
```

@@

This paper did not examine or compare ridership data between conventional and equity stations. Therefore, further research is needed to determine whether this encouraged more cycling for low-income individuals living near equity stations. Other studies have specifically looked at differences in trip type, frequency, or length among users from disadvantaged neighbourhoods [@caspiBikesharingPhiladelphiaLowerincome2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], but our analysis is limited by the lack of accessible data to conduct similar analyses for Hamilton Bike Share.

@@

The addition of specific equity stations to the public bicycle share program in Hamilton, Ontario had the net effect of increasing accessibility and reducing both horizontal and vertical inequities. In particular, accessibility improved the most for the second 20% income groups at all thresholds, but the gains were only modest. Congestion effects were observed at higher thresholds, with accessibility decreasing significantly once the catchment area is increased to 10 minutes walking. However, We identified specific areas that have both low accessibility and low income which would benefit from an increased supply of bicycles. These are ideal candidates for new equity stations. In this way, our paper has made contributions in a positive way by applying an intuitive and useful approach to measure accessibility to a PBSP, and in a normative way by serving to inform future investments in cycle infrastructure for Hamilton Bike Share.

>>>>

This paper uses the city of Hamilton, located in Ontario, Canada, as a case study. The city launched a public bicycle share program, Hamilton Bike Share, in March 2015 with 115 stations and 750 bicycles [@hamiltonHamiltonBikeShare2015]. Before June 2020, the program was known as SoBi Hamilton. Stations are spaced between 300 and 600 metres apart [@scottRouteChoiceBike2021]. The core service area spans 35 km^2 of the city and roughly 138,000 can reach a bike share hub within 30 minutes of walking [see Figure \ref{fig:hamilton-and-sobi-service-area}]. This represents roughly one fifth of the population in Hamilton, according to the 2016 Canadian Census. The program was enthusiastically welcomed in the city - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015]. In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, Everyone Rides Initiative (ERI), to remove barriers that may prevent individuals from accessing bike share in Hamilton. An additional 75 bicycles and 15 stations were added to the program, which expanded it to more disadvantaged areas in the core service area [see Figure 2]. The program also offers subsidized memberships to individuals who identify as low income. A comparable program can be found in Philadelphia [see @caspiBikesharingPhiladelphiaLowerincome2019]. As of June 2020, the bike share program has 900 bikes and 130 stations [see Figure \ref{fig:sobi-stations-in-hamilton}], and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].

@@

```{r hamilton-and-sobi-service-area, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "The core service area of Hamilton Bike Share is outlined in black. Hamilton Census Metropolitan Area (CMA) is shown in grey."}
ggplot() + 
  geom_sf(data = hamilton_da_2016,
          color = NA,
          fill = "gray") + 
  geom_sf(data = sobi_service,
          fill = NA,
          colour = "black") + 
  theme(panel.background = element_rect(fill = "white"))
```

```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "100%", fig.cap = "The spatial distribution of bike share stations in Hamilton, Ontario. The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in dark green. Hamilton CMA is shown in grey."}
ggplot() + 
  geom_sf(data = cropped_da,
          color = NA,
          fill = "gray") +
  geom_sf(data = sobi_service, 
          fill = NA,
          color = "black") +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          size = 1,
          color = "dark green") +
  geom_sf(data = sobi_hubs %>% 
            dplyr::filter(hub_type == "Conventional" & 
                            hub_status == "Active"), # Remove deactivated hubs
          aes(size = RACKS_AMOU),
          shape = 21,
          fill = "blue") +
  geom_sf(data =  sobi_hubs %>% 
            dplyr::filter(hub_type == "ERI"), 
          aes(size = RACKS_AMOU), 
          shape = 21,
          fill = "green") +
  scale_size(name = "Number of racks", range = c(0.5, 4)) +
  theme(panel.background = element_rect(fill = "white"))
```

@@

Following the method first presented by Tobler [-@toblerSmoothPycnophylacticInterpolation1979], we used pycnophylactic interpolation to disaggregate population from each dissemination area to smaller polygons that are 50 by 50 metres. This method involves smoothing out the population from each dissemination area while preserving total volume [see Figure \ref{fig:da-population}]. Since pycnophylactic interpolation occurs at such a micro scale, we had to ensure that population numbers were not reallocated or smoothed out to areas where people do not live in Hamilton (i.e., where schools or parks are located). To do so, we retrieved shapefiles for various geographic features [see Table \ref{tab:data-features}] from Open Hamilton. Next, we extracted these features from the PBSP core service area and used pycnophylactic interpolation to disaggregate and reallocate population [see Figure \ref{fig:interpolated-population}].

@@

kable(data_tbl,
      "latex",
      booktabs = TRUE,
      caption = "\\label{tab:landscape-features}Landscape features extracted from the Hamilton Bike Share core service area before population is interpolated.") %>%
  kable_paper(full_width = F) %>%
  column_spec(1, bold = T, border_right = T) %>%
  column_spec(2, width = "30em") %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

```{r da-population, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Population in all dissemination areas (outlined in white) in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the service area."}
ggplot() +
  geom_sf(data = hamilton_da_2016 %>%
            dplyr::filter(sobi_service == "Yes"),
          aes(fill = population),
          color = "white") +
  geom_sf(data = sobi_service,
          fill = NA) +
  scale_fill_distiller(palette = "YlOrRd", 
                       direction = 1) +
  theme(panel.background = element_rect(fill = "white"))
```

```{r interpolated-population, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Interpolated population in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue). The downtown area is outlined in dark green. Geographic features have been extracted. Dissemination area polygons are outlined in grey."}
ggplot() + 
  geom_sf(data = cropped_da,
          fill = "gray",
          color = NA) + 
  geom_sf(data = population_50x50,
          aes(color = population)) +
  scale_color_distiller(palette = "YlOrRd", 
                        direction = 1) +
  geom_sf(data = sobi_service,
          fill = NA,
          colour = "black") +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  theme(panel.background = element_rect(fill = "white"))
```

@@

@@

Consensus regarding the distance that individuals are willing to walk to access a PBSP is lacking, therefore the literature on walking behaviour was consulted to determine the thresholds for the sensitivity analysis. Previous studies have found that living within 250 metres [@fullerUseNewPublic2011] and 300 metres [@kabraBikeShareSystemsAccessibility2020] is correlated with bike share use, while other research has found that walking trips are less than 600 metres and rarely more than 1200 metres [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 metres [@larsenQuarterMileReexamining2010]. We found that accessibility increases with a threshold between two and four minutes, but is then maximized at 5 minutes. At some stations, Hamilton Bike Share will depict a map to show the user the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are anticipated to walk. Accessibility decreases significantly after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each bike share station.

@@

@@

```{r accessibility-sensibility, eval=FALSE, include=FALSE}
# Testing different thresholds to check how accessibility changes
accessibility_test <- data.frame(threshold = c(3:15), 
                                 accessibility = numeric(13))
for(i in 1:13){
  results_test <- ttm_walk %>%
    dplyr::filter(hub_type == "Conventional" &
                    hub_status == "Active") %>%
    b2sfca(threshold = i +2 )
  accessibility_test$accessibility[i] <- sum(results_test$accessibility$accessibility)
}

@@

At the minimum threshold, we find that there are 25.2 bicycle racks per person in the original system.  With the addition of equity stations, there are now 25.4 bicycles per person. Figure \ref{fig:figure-6} presents a comparison of accessibility between the systems. Accessibility is fairly uniform, with the exception of two small areas near the university and the waterfront park where accessibility is slightly higher.

@@

At the average threshold, we find that there are 68.6 bicycle racks per person in the original system.  With the addition of equity stations, there are now 68.8 bicycles per person. Figure \ref{fig:figure-7} presents a comparison of accessibility between the systems. Again, accessibility is fairly uniform, with the exception of one very small area near the waterfront park where accessibility is higher.

@@

At the maximum threshold, we find that there are 3.61 bicycle racks per person in the original system.  With the addition of equity stations, there are now 3.74 bicycles per person. Figure \ref{fig:figure-8} presents a comparison of accessibility between the systems. We begin to see differences in accessibility across the service area, with users near the university and its adjacent neighbourhoods, as well as neighbourhoods north of the downtown area, have slightly higher accessibility. While the differences are modest, they are more apparent at this threshold than at shorter walking distances.

@@

At the extreme threshold, we find that there are 2.44 bicycle racks per person in the original system.  With the addition of equity stations, there are now 2.55 bicycles per person. Figure \ref{fig:figure-9} presents a comparison of accessibility between the systems. Users near the university and the neighbourhoods north of the downtown area (outlined in green) have the highest accessibility, followed by those who live in the city's downtown area. Accessibility in the east end of the core service area remains low.

@@

Approximately 138,000 people live within Hamilton Bike Share's core service area or within a 30 minute walk to the service area. We find that over 118,000 people can access a bicycle share station within a 15 minute walk, which represents roughly 85% of the total population in the core service area [see Table \ref{tab:accessibility-income}]. While previous research found that neighbourhoods with more disadvantage are better serviced by Hamilton Bike Share [@hosfordWhoArePublic2018], the authors used the Pampalon Deprivation Index from 2011 as a measure of socioeconomic status for dissemination areas across the city not just within the core service area. Instead, we use median total income for each dissemination area in the core service area to examine whether disparities in accessibility exist between income groups for those who live in or near the core service area. We divide income by quantiles: bottom 20%, second 20%, third 20%, fourth 20%, and top 20%. One of the unique properties of the balanced floating catchment area method is that data can be reaggregated from small to larger polygons while preserving the total population and supply at each station. This avoids demand and supply inflation, and also enables us to present findings in a way that is easier to interpret (i.e., across dissemination areas instead of 50 by 50 metre population cells).

Panels 1[\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps with the combined spatial distribution of accessibility and median total income per household for the different thresholds. Our analysis demonstrates that the extreme threshold of fifteen minutes allows the most people to access a station. We find that stations added to Hamilton's public bicycle share program to increase equity for disadvantaged neighbourhoods achieved their goal by increasing accessibility, albeit only modestly [see Table \ref{tab:accessibility-income}]. By implementing equity stations in more areas with lower median total income in Hamilton, the PBSP has achieved greater horizontal equity by extending the spatial distribution of bicycles across the city. This is particularly evident at the minimum and average thresholds of three and five minutes, respectively. The largest gains were made for dissemination areas in the second 20%, where an additional 3,073 and 5,395 people could reach a bicycle share station within three and five minutes walk, respectively, after the addition of the equity station.

However, we found that dissemination areas with the lowest total household income do not have much greater access to the program [see Table \ref{tab:table-3}]. Income disparities still persist, however only at certain thresholds. With and without equity stations, people in the top 20% of income have the highest level of accessibility at a threshold of ten and fifteen minutes. Although dissemination areas in the second 20% have the highest level of accessibility by a significant amount at lower thresholds, the bottom 20%, who may benefit the most from Hamilton Bike Share's equity initiatives, have the lowest accessibility at three minutes threshold and the second lowest accessibility at all other thresholds. Panels 1[\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] reveal specific areas where accessibility is low or zero, which are good candidates for additional equity stations.

On the whole, we find that areas that have less than their equitable share of the level of service are some of the most socioeconomically disadvantaged areas in the city. While the addition of equity stations increases accessibility for all income groups at all thresholds, they did not increase accessibility significantly for any single income group. This finding aligns with other studies that have found disparities in station or bike location along income lines in New York City [@babagoliExploringHealthSpatial2019], Tampa [@chenExploringEquityPerformance2019], and Seattle [@mooneyFreedomStationSpatial2019], where people who have greater advantage have better access to PBSPs.

@@

kable(accessibility_income,
      "latex",
      booktabs = TRUE,
      col.names = c("Income Quintile",
                    "Accessibility",
                    "Population",
                    "Accessibility",
                    "Population",
                    "Accessibility",
                    "Population"),
      align = c("l", "c", "c", "c", "c", "c", "c"),
      caption = "\\label{tab:accessibility-by-income}Accessibility and population serviced by income quintile and between systems (with and without equity stations).") %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  pack_rows("Threshold - 3 minutes", 1, 5, 
            label_row_css = "background-color: #8f9498; color: #fff;") %>%
  pack_rows("Threshold - 5 minutes", 6, 10, 
            label_row_css = "background-color: #8f9498; color: #fff;") %>%
  pack_rows("Threshold - 10 minutes", 11, 15, 
            label_row_css = "background-color: #8f9498; color: #fff;") %>%
  pack_rows("Threshold - 15 minutes", 16, 20, 
            label_row_css = "background-color: #8f9498; color: #fff;") %>%
  add_header_above(c(" " = 1, "Without Equity Stations" = 2, "With Equity Stations" = 2, "Difference" = 2)) %>%
  #column_spec(2, color = "white",
  #background = spec_color(accessibility_income$accessibility[1:40], end = 1)) %>%
  footnote(general = " ",
           alphabet = c("With equity stations = Hamilton Bike Share current system (118 conventional stations, 12 equity stations)", "Without equity stations = Hamilton Bike Share original system (118 conventional stations, no equity stations)"))
```

@@

This paper did not examine or compare ridership data between conventional and equity stations. Therefore, further research is needed to determine whether the addition of equity stations encouraged more cycling for low-income individuals living near them. Other studies have specifically looked at differences in trip type, frequency, or length among users from disadvantaged neighbourhoods [@caspiBikesharingPhiladelphiaLowerincome2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], but our analysis is limited by the lack of accessible route or user data to conduct similar analyses for Hamilton Bike Share.

@@

The addition of specific equity stations to the public bicycle share program in Hamilton, Ontario had the net effect of increasing accessibility and reducing both horizontal and vertical inequities. In particular, accessibility improved the most for the second 20% median total income groups at all thresholds, but the gains were only modest for all income groups. Dissemination areas with the bottom 20% had the lowest accessibility at three minutes, and second lowest level of accessibility at five, ten, and fifteen minutes. Congestion effects were observed at higher thresholds, with accessibility decreasing significantly once the catchment area is increased to 10 minutes walking. Based on our analysis, we identified specific areas that have both low accessibility and low income which would benefit from an increased supply of bicycles. These are ideal candidates for new equity stations. In this way, our paper has made contributions in a positive way by applying an intuitive and useful approach to measure accessibility to a PBSP, and in a normative way by serving to inform future investments in cycle infrastructure for Hamilton Bike Share.
